Symbolic Weighted Language Models and Parsing over Infinite Alphabets

### 2021-05-06 main.tex
details section VPA

\section{SW Automata and Transducers}
\label{section:transducer}
\label{section:SWA}
\label{section:SWT}

@@

We give now a formal definition of these transitions of the automaton $A$
in term of a weight value
computed by an intermediate function $\weight_A$.
In the case of a pushdown automaton, a configuration is composed 
of a state $q \in Q$ and a stack content $\theta \in \Theta^*$, where $\Theta = \Omegac \times P$.
Therefore, $\weight_A$ is a function from 
$Q \times \Theta^* \times \Omega^* \times Q \times \Theta^*$ into $\Semiring$.
\[
\begin{array}{rcl}
\weight_A\bigl(\config{q}{\theta}, a\, u, \config{q'}{\theta'}\bigr) & = & 
 {\displaystyle\bigoplus_{q'' \in Q}} \weii(q, a, q'') 
 \otimes \weight_A\bigl(\config{q''}{\theta}, u, \config{q'}{\theta'}\bigr)\\
%
\weight_A\bigl(\config{q}{\theta}, {\call{c}}\, u, \config{q'}{\theta'}\bigr) & = & 
 {\displaystyle\bigoplus_{\begin{array}{c}
                          \scriptstyle q'' \in Q\\[-2pt]
                          \scriptstyle p \in P
                          \end{array}}}
 \weic\bigl(q, {\call{c}}, q'', p\bigr) 
 \otimes \weight_A\bigl(\config{q''}{{\call{c}}\, p\cdot \theta}, u, \config{q'}{\theta'}\bigr)\\[1mm]
%
\weight_A\bigl(\config{q}{{\call{c}}\, p\cdot \theta}, {\return{r}}\, u, \config{q'}{\theta'}\bigr) & = & 
 {\displaystyle\bigoplus_{q'' \in Q}} 
 \weir\bigl(q, {\call{c}}, p, {\return{r}}, q''\bigr) 
 \otimes \weight_A\bigl(\config{q''}{\theta}, u, \config{q'}{\theta'}\bigr)\\
%
\weight_A\bigl(\config{q}{\bot}, {\return{r}}\, u, \config{q'}{\theta'}\bigr) & = & 
 {\displaystyle\bigoplus_{q'' \in Q}} \weie(q, {\return{r}}, q'') 
 \otimes \weight_A\bigl(\config{q''}{\bot}, u, \config{q'}{\theta'}\bigr)\\
\end{array}
\]
where $\bot$ denotes the empty stack and ${\call{c}}\, p\cdot \theta$ 
denotes a stack where the pair made of ${\call{c}}$ and $p$ is the top symbol and $\theta$
is the rest of stack.

@@

\subsection{Properties}
The class of \SWVPA is closed under the binary operators of the underlying semiring.
%
\begin{proposition}
Let $A_1$ and $A_2$ be two (\SWVPA)
over the same $\Omega$ and $\Semiring$.
There exists two $\SWVPA$ $A_1 \oplus A_2$ and $A_1 \otimes A_2$, 
effectively constructible, 
such that for all $s \in \Omega^*$, 
$(A_1 \oplus A_2)(s) = A_1(s) \oplus A_2(s)$ and 
$(A_1 \otimes A_2)(s) = A_1(s) \otimes A_2(s)$.
\end{proposition}
The construction is essentially the same 
as the one of~\cite{dAntonyAlur14SVPDA}, in the case of the Boolean semiring.

@@

\noindent
More precisely, 
let $\top$ be a fresh stack symbol which does not belong to $\Theta$,
and for every two states $q, q' \in Q$ and $\sigma \in \{ \bot, \top \}$, 
let 
\[ 
  d_0(q, \sigma, q') = \bigoplus_{s\in \Omega^*} 
  \textstyle
  \weight_A\bigl(\config{q}{\sigma}, s, \config{q'}{\sigma}\bigr). 
\]
Since $\Semiring$ is complete, this infinite sum is well defined, 
and since $\leq_\oplus$ is assumed total, it is the minimum in $\Omega^*$
of $s \mapsto \weight_A(\config{q}{\sigma}, s, \config{q'}{\sigma})$ \wrt this ordering.
%
When $\sigma = \bot$, $d_0(q, \sigma, q')$ is the central 
expression in a term of the definition (\ref{eq:weightA}) of $A(s_0)$ for the minimum $s_0$
(for the above function).
When $\sigma = \bot$, intuitively, it is the minimum weight of a computation of $A$
starting in state $q$ with a stack $\theta \in \Theta^*$ (possibly empty), 
and ending in state $q$ with the same stack $\theta$, such that moreover 
the computation does not touch a symbol of $\theta$. 
That means that during the computation, 
$A$ may apply the first case of in the definition of $\weight_A$ (internal symbol),
as well as the second case, to can push call symbols on the top of $\theta$, 
and may pop these symbols with the third case (return symbol). 
However, it cannot apply one of the two last cases (return symbol and empty stack)
when the current stack is $\theta$.
%pop symbols in $\theta$.
% Note that having a stack reduced to such a symbol makes impossible the application of the 
% two last cases in the definition of $\weight_A$ (return symbol and empty stack). 
% However, it is possible to apply the two first cases 
% (internal symbol or call symbol, with a push on the top of $\top$).

>>>>

\section{SW Automata and Transducers}
\label{section:transducer}\label{sec:transducer}
\label{section:SWA}\label{sec:SWA}
\label{section:SWT}\label{sec:SWT}

@@

We give now a formal definition of these transitions of the automaton $A$
in term of a weight value
computed by an intermediate function $\weight_A$, like in Section~\ref{sec:SWT}.
In the case of a pushdown automaton, a configuration is composed 
of a state $q \in Q$ and a stack content $\gamma \in \Gamma^*$, where $\Gamma = \Omegac \times P$.
Therefore, $\weight_A$ is a function from 
$Q \times \Gamma^* \times \Omega^* \times Q \times \Gamma^*$ into $\Semiring$.
\[
\begin{array}{rcl}
\weight_A\bigl(\config{q}{\gamma}, a\, u, \config{q'}{\gamma'}\bigr) & = & 
 {\displaystyle\bigoplus_{q'' \in Q}} \weii(q, a, q'') 
 \otimes \weight_A\bigl(\config{q''}{\gamma}, u, \config{q'}{\gamma'}\bigr)\\
%
\weight_A\bigl(\config{q}{\gamma}, {\call{c}}\, u, \config{q'}{\gamma'}\bigr) & = & 
 {\displaystyle\bigoplus_{\begin{array}{c}
                          \scriptstyle q'' \in Q\\[-2pt]
                          \scriptstyle p \in P
                          \end{array}}}
 \weic\bigl(q, {\call{c}}, q'', p\bigr) 
 \otimes \weight_A\bigl(\config{q''}{{\call{c}}\, p\cdot \gamma}, u, \config{q'}{\gamma'}\bigr)\\[1mm]
%
\weight_A\bigl(\config{q}{{\call{c}}\, p\cdot \gamma}, {\return{r}}\, u, \config{q'}{\gamma'}\bigr) & = & 
 {\displaystyle\bigoplus_{q'' \in Q}} 
 \weir\bigl(q, {\call{c}}, p, {\return{r}}, q''\bigr) 
 \otimes \weight_A\bigl(\config{q''}{\gamma}, u, \config{q'}{\gamma'}\bigr)\\
%
\weight_A\bigl(\config{q}{\bot}, {\return{r}}\, u, \config{q'}{\gamma'}\bigr) & = & 
 {\displaystyle\bigoplus_{q'' \in Q}} \weie(q, {\return{r}}, q'') 
 \otimes \weight_A\bigl(\config{q''}{\bot}, u, \config{q'}{\gamma'}\bigr)\\
\end{array}
\]
where $\bot$ denotes the empty stack and ${\call{c}}\, p\cdot \gamma$ 
denotes a stack where the pair made of ${\call{c}} \in \Omegac$ and $p \in P$ is the top symbol 
and $\gamma$ is the rest of stack.

@@

\subsection{Properties}
The class of \SWVPA is closed under the binary operators of the underlying semiring.
%
\begin{proposition}
Let $A_1$ and $A_2$ be two (\SWVPA)
over the same $\Omega$ and $\Semiring$.
There exists two $\SWVPA$ $A_1 \oplus A_2$ and $A_1 \otimes A_2$, 
effectively constructible, 
such that for all $s \in \Omega^*$, 
$(A_1 \oplus A_2)(s) = A_1(s) \oplus A_2(s)$ and 
$(A_1 \otimes A_2)(s) = A_1(s) \otimes A_2(s)$.
\end{proposition}
The construction is essentially the same 
as in the case of the Boolean semiring~\cite{dAntonyAlur14SVPDA}.

@@

\noindent
More precisely, 
let $\top$ be a fresh stack symbol which does not belong to $\Gamma$,
and for every two states $q, q' \in Q$ and $\sigma \in \{ \bot, \top \}$, 
let 
\[ 
  d_0(q, \sigma, q') = \bigoplus_{s\in \Omega^*} 
  \textstyle
  \weight_A\bigl(\config{q}{\sigma}, s, \config{q'}{\sigma}\bigr). 
\]
Since $\Semiring$ is complete, this infinite sum is well defined, 
and since $\leq_\oplus$ is assumed total, it is the minimum in $\Omega^*$
of $s \mapsto \weight_A(\config{q}{\sigma}, s, \config{q'}{\sigma})$ \wrt this ordering.
%
When $\sigma = \bot$, $d_0(q, \sigma, q')$ is the central 
expression in a term of the definition (\ref{eq:weightA}) of $A(s_0)$ for the minimum $s_0$
(for the above function).
When $\sigma = \bot$, intuitively, it is the minimum weight of a computation of $A$
starting in state $q$ with a stack $\gamma \in \Gamma^*$ (possibly empty), 
and ending in state $q$ with the same stack $\gamma$, such that moreover 
the computation does not touch a symbol of $\gamma$. 
That means that during the computation, 
$A$ may apply the first case of in the definition of $\weight_A$ (internal symbol),
as well as the second case, to can push call symbols on the top of $\gamma$, 
and may pop these symbols with the third case (return symbol). 
However, it cannot apply one of the two last cases (return symbol and empty stack)
when the current stack is $\gamma$.
%pop symbols in $\gamma$.
% Note that having a stack reduced to such a symbol makes impossible the application of the 
% two last cases in the definition of $\weight_A$ (return symbol and empty stack). 
% However, it is possible to apply the two first cases 
% (internal symbol or call symbol, with a push on the top of $\top$).
